Illustrating Model Properties

### 2024-06-05 model.qmd
switch to marginals over unknown combos

```{python}
#| label: fig-model-prevalences-overall
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
import matplotlib.pyplot as plt
from matplotlib import gridspec
import numpy as np
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
from scripts import paths, shared
from scripts.marginalize_midext import compute_possible_observed_prevalences, SummedBetaPosterior

@@

for i, t_stage in enumerate(["early", "late"]):
  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
  for j, midext in enumerate([False, True]):
    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
    axes[0,j].set_xticks([])
    axes[-1,j].set_xlabel("prevalence [%]")
    axes[i,j].set_yticks([])
    content = []
    for lnl, color in zip(
      ["II", "III", "IV"],
      [COLORS["blue"], COLORS["orange"], COLORS["green"]],
    ):
      scenario = Scenario(
        t_stages=[t_stage],
        midext=midext,
        diagnosis={"contra": {"max_llh": {lnl: True}}}
      )
      content.append(
        Histogram.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label="prediction",
        )
      )
      content.append(
        BetaPosterior.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
        )
      )
      num_success, num_total = compute_possible_observed_prevalences(data, scenario).T
      content.append(
        SummedBetaPosterior(
          num_success=num_success,
          num_total=num_total,
          kwargs={"label": None, "linestyle": "--"},
        )
      )
    draw(
      axes=axes[i,j],
      contents=content,
      xlims=(0., 15. if not midext else 45.),
    )

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.57, 0.55), (0.48, 0.99), (0.15, 0.99)],
)
split_legends(
  axes=axes[0,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.67, 0.99), (0.27, 0.6), (0.1, 0.99)],
)
split_legends(
  axes=axes[1,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.27, 0.6), (0.24, 0.99)],
)
split_legends(
  axes=axes[1,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.4, 0.6), (0.16, 0.99)],
)

@@

```{python}
#| label: fig-model-prevalences-with-ipsi
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
import matplotlib.pyplot as plt
from matplotlib import gridspec
import numpy as np
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
from scripts import paths, shared
from scripts.marginalize_midext import (
  compute_possible_observed_prevalences,
  SummedBetaPosterior,
)

@@

for i, t_stage in enumerate(["early", "late"]):
  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
  for j, midext in enumerate([False, True]):
    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
    axes[0,j].set_xticks([])
    axes[-1,j].set_xlabel("prevalence [%]")
    axes[i,j].set_yticks([])
    content = []
    for ipsi, color in zip(
      [{"I": False, "II": False, "III": False, "IV": False, "V": False}, {"III": True}],
      [COLORS["blue"], COLORS["orange"]],
    ):
      scenario = Scenario(
        t_stages=[t_stage],
        midext=midext,
        diagnosis={
          "ipsi": {"max_llh": ipsi},
          "contra": {"max_llh": {"II": True}},
        }
      )
      content.append(
        Histogram.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label="prediction",
        )
      )
      content.append(
        BetaPosterior.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
        )
      )
      num_success, num_total = compute_possible_observed_prevalences(data, scenario).T
      content.append(
          SummedBetaPosterior(
          num_success=num_success,
          num_total=num_total,
          kwargs={"linestyle": "--", "label": None},
        )
      )
    draw(
      axes=axes[i,j],
      contents=content,
      xlims=(0., 30. if midext else 10.),
    )

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.45, 0.55)],
)
split_legends(
  axes=axes[0,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[1,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.2, 0.99), (0.55, 0.99)],
)
split_legends(
  axes=axes[1,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.2, 0.99), (0.65, 0.99)],
)

@@

for t_stage in ["early", "late"]:
  has_t_stage = data.ly.t_stage.isin([t_stage])
  scenario = Scenario(
    t_stages=[t_stage],
    midext=True,
    diagnosis={"ipsi": {"max_llh": {"II": None}}},
  )
  unknowns = (has_unknown_midext & has_t_stage).sum()
  color = next(colors)
  content.append(
    Histogram.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
      dataname=t_stage,
      color=color,
      label="prediction"
    )
  )
  beta_post = BetaPosterior.from_hdf5(
    filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
    dataname=t_stage,
    color=color,
  )
  content.append(beta_post)
  combs = possible_combinations(
    matches=unknowns,
    max_total=unknowns,
  )
  content.append(
    SummedBetaPosterior(
      num_success=combs.matches.values + beta_post.num_success,
      num_total=combs.total.values + beta_post.num_total,
      permutations=combs.permutations.values,
      kwargs={"linestyle": "--", "label": None, "color": "red"},
    )
  )
  print(f"{unknowns = }, {beta_post.num_success = }, {beta_post.num_total = }")

>>>>

```{python}
#| label: fig-model-prevalences-overall
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
import matplotlib.pyplot as plt
from matplotlib import gridspec
import numpy as np
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
from scripts import paths, shared
from scripts.marginalize_midext import possible_prevalence_combinations, SummedBetaPosterior

@@

for i, t_stage in enumerate(["early", "late"]):
  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
  for j, midext in enumerate([False, True]):
    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
    axes[0,j].set_xticks([])
    axes[-1,j].set_xlabel("prevalence [%]")
    axes[i,j].set_yticks([])
    content = []
    for lnl, color in zip(
      ["II", "III", "IV"],
      [COLORS["blue"], COLORS["orange"], COLORS["green"]],
    ):
      scenario = Scenario(
        t_stages=[t_stage],
        midext=midext,
        diagnosis={"contra": {"max_llh": {lnl: True}}}
      )
      content.append(
        Histogram.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label="prediction",
        )
      )
      # content.append(
      #   BetaPosterior.from_hdf5(
      #     filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
      #     dataname=scenario.md5_hash("prevalences"),
      #     color=color,
      #   )
      # )
      combs = possible_prevalence_combinations(data, scenario)
      content.append(
        SummedBetaPosterior(
          num_success=combs.matches.values,
          num_total=combs.total.values,
          permutations=combs.permutations.values,
          kwargs={"color": color},
        )
      )
    draw(
      axes=axes[i,j],
      contents=content,
      xlims=(0., 15. if not midext else 45.),
    )

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.53, 0.6), (0.55, 0.99), (0.12, 0.99)],
)
split_legends(
  axes=axes[0,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.55, 0.99), (0.27, 0.6), (0.1, 0.99)],
)
axes[1,0].set_ylim(top=1.4)
split_legends(
  axes=axes[1,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.54, 0.99), (0.24, 0.6), (0.11, 0.99)],
)
split_legends(
  axes=axes[1,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.54, 0.99), (0.4, 0.6), (0.12, 0.99)],
)

@@

```{python}
#| label: fig-model-prevalences-with-ipsi
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
import matplotlib.pyplot as plt
from matplotlib import gridspec
import numpy as np
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
from scripts import paths, shared
from scripts.marginalize_midext import (
  possible_prevalence_combinations,
  SummedBetaPosterior,
)

@@

for i, t_stage in enumerate(["early", "late"]):
  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
  for j, midext in enumerate([False, True]):
    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
    axes[0,j].set_xticks([])
    axes[-1,j].set_xlabel("prevalence [%]")
    axes[i,j].set_yticks([])
    content = []
    for ipsi, color in zip(
      [{"I": False, "II": False, "III": False, "IV": False, "V": False}, {"III": True}],
      [COLORS["blue"], COLORS["orange"]],
    ):
      scenario = Scenario(
        t_stages=[t_stage],
        midext=midext,
        diagnosis={
          "ipsi": {"max_llh": ipsi},
          "contra": {"max_llh": {"II": True}},
        }
      )
      content.append(
        Histogram.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label="prediction",
        )
      )
      # content.append(
      #   BetaPosterior.from_hdf5(
      #     filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
      #     dataname=scenario.md5_hash("prevalences"),
      #     color=color,
      #   )
      # )
      combs = possible_prevalence_combinations(data, scenario)
      content.append(
        SummedBetaPosterior(
          num_success=combs.matches.values,
          num_total=combs.total.values,
          permutations=combs.permutations.values,
          kwargs={"color": color},
        )
      )
    draw(
      axes=axes[i,j],
      contents=content,
      xlims=(0., 30. if midext else 10.),
    )

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[0,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[1,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.16, 0.99), (0.53, 0.55)],
)
split_legends(
  axes=axes[1,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.13, 0.55), (0.5, 0.99)],
)

@@

for t_stage in ["early", "late"]:
  has_t_stage = data.ly.t_stage.isin([t_stage])
  scenario = Scenario(
    t_stages=[t_stage],
    midext=True,
    diagnosis={"ipsi": {"max_llh": {"II": None}}},
  )
  unknowns = (has_unknown_midext & has_t_stage).sum()
  color = next(colors)
  content.append(
    Histogram.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
      dataname=t_stage,
      color=color,
      label="prediction"
    )
  )
  content.append(
    BetaPosterior.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
      dataname=t_stage,
      color=color,
    )
  )
